Modelling $q_{95}$ Windows for the Suppression of Edge Localized Modes by Resonant Magnetic Perturbations in the DIII-D Tokamak

### 2020-07-11 ExperimentalData/DIIID/145380/Paper/paper.tex
First circulated version of paper
Version of paper submitted to GA

\begin{abstract}
A toroidal asymptotic matching model of the response of a tokamak plasma to a static resonant magnetic perturbation (RMP) is used to simulate the $n=3$ RMP-induced 
edge-localized-mode (ELM)-suppression
windows in $q_{95}$ that are evident when the plasma current is slowly ramped in DIII-D discharge \#145380. All quantities employed in the simulation are
derived from experimental measurements, apart from the neutral particle data. 
Three cases are considered.
In the first case, the natural frequencies of tearing modes resonant in the plasma are 
determined by the ion flows at the corresponding rational surfaces, which is the prediction of nonlinear theory.
In the second case, the natural frequencies are 
determined by the guiding center flows at the rational surfaces.
 In the third case, the natural frequencies   are 
determined by the electron flows at the  rational surfaces, which is the prediction of linear theory.
The second case gives  the best agreement between the simulations and the  experimental observations.   The third case gives the worst agreement between the simulations 
and the  observations. The first case only leads to partial agreement between  the  simulations and the  observations. However, this lack of complete agreement
may be a consequence of using an inaccurate guess for the neutral particle distribution in the pedestal.  
\end{abstract}

@@

Figure~\ref{fig7} shows the widths of the regions of density and electron temperature flattening associated with the $n=3$ magnetic
island chains driven by the RMP in the pedestal region as functions of time. (See Sect.~III.F in Ref.~\onlinecite{rf2}.) (The particle and energy perpendicular diffusivities are assumed to be the
same as the momentum perpendicular diffusivity.) It can be seen that the driven magnetic island chains are generally too narrow
to significantly flatten the density and electron temperature profiles,\cite{flat}  implying that there is actually very little degradation of the pedestal energy and particle confinement
due to the applied RMP. 
However, in those situations in which the shielding of driven
magnetic reconnection due to plasma flow breaks down (i.e., the true natural frequency becomes zero), locked magnetic island chains are generated that are
 wide enough to flatten both the density and the electron temperature profiles.

 Ideally, in order to account for the
ELM-suppression windows in DIII-D discharge \#145380, according to the standard
hypothesis of how RMP-induced ELM-suppression operates, we require the major regions of density and temperature
flattening to correspond to the intersections of the yellow vertical bands and the horizontal dotted lines in Fig.~\ref{fig7}. If this is the case then
the density and temperature profiles are locally flattened at the top of the pedestal (which corresponds to
the horizontal dotted lines) during the ELM-suppression windows (which correspond to the yellow vertical bands). It can be seen
that the region of density and temperature flattening associated with the $m=11$ (brown) tearing mode driven by the RMP lines up fairly well with the
first ELM-suppression window. The region of density and temperature flattening associated with the $m=10$ (magenta) tearing mode 
lines up somewhat less well with the second ELM-suppression window. To be more exact, the region is located at the top of the pedestal,
but occurs slightly too early. However, the 
region of density and temperature flattening associated with the $m=9$ (cyan) tearing mode 
lines up very poorly with the third ELM-suppression window. To be more exact, the region is not really located at the top of the pedestal, and occurs
significantly too early. The ultimate reason for this poor alignment is that the major zero-crossing of the $m=9$ natural
frequency in the upper panel of Fig.~\ref{fig5} does not take place within the third ELM-suppression window (unlike the
major zero-crossings of the $m=11$ and $m=10$ natural frequencies, which do take place within the first and the second
ELM-suppression windows, respectively).

@@

\section*{Acknowledgements}
This research was funded by the U.S.\ Department of Energy under contract DE-FG02-04ER-54742.
The author would like to thank Q.M.~Hu and R.~Nazikian for providing the experimental data used in this paper.

@@

Figure~\ref{fig14} also sketches the
typical plasma current and vorticity patterns in the nonlinear regime. Observe that there is now a considerable distinction between an X-point and an O-point. The reason for this is that the island is sufficiently wide that the plasma cannot diffuse across it resistively rapidly enough to
avoid being trapped inside the magnetic separatrix. Indeed, both the regions inside and outside the magnetic separatrix
are governed by a combination of flux-freezing and perturbed force balance, whereas the resistive
layer, within which the plasma can easily slip through the magnetic field,  is diverted onto the magnetic separatrix. Moreover, as
the island grows, the width of the separatrix layer becomes
increasing negligible with respect to the island width. It
turns out that the separatrix layer does not significantly affect the evolution of a nonlinear magnetic island chain. Indeed,
the celebrated Rutherford island width evolution equation,
which governs the evolution of the island width, is simply a
$1\tfrac{1}{2}$-D evolution equation for a helical magnetic
equilibrium localized around the rational surface.\cite{ruth}
The Alfv\'{e}n time plays no role in this evolution (any more
than it plays a role in the $1\tfrac{1}{2}$-D evolution of
the global magnetic field). Likewise, the equation that
determines the phase velocity of a nonlinear magnetic island chain (see Ref.~\onlinecite{nl2}, and references therein) is essentially an
expression of the $1\tfrac{1}{2}$-D evolution of the plasma
flow profiles across the island region. As before, the  Alfv\'{e}n time plays no role in this evolution (any more
than it plays a role in the $1\tfrac{1}{2}$-D evolution of
the global plasma flow profiles).

@@

\begin{figure}
\includegraphics[height=7in]{fig3.pdf}
\caption{Top Panel: The red, green, blue, and cyan curves show the electron number density ($10^{19}\,{\rm m}^{-3}$),
electron temperature (keV), (thermal) ion temperature (keV), and
 C-VI ion number density  ($10^{18}\,{\rm m}^{-3}$)  profiles, respectively,  in  DIII-D discharge \#145380 at time $t=2500$ ms. Middle Panel:  
 ${\bf E}\times {\bf B}$ frequency profile in  DIII-D discharge \#145380 at time $t=2500$. Bottom Panel: Perpendicular momentum diffusivity 
profile in  DIII-D discharge \#145380 at time $t=2500$ ms.
The   common vertical dotted lines indicate the location of the top
of the pedestal, ${\mit\Psi}_N=0.925$.} \label{fig3}
\end{figure}

>>>>

\begin{abstract}
A toroidal asymptotic matching model of the response of a tokamak plasma to a static resonant magnetic perturbation (RMP) is used to simulate the $n=3$ RMP-induced 
edge-localized-mode (ELM)-suppression
windows in $q_{95}$ that are evident when the plasma current is slowly ramped in DIII-D discharge \#145380. All quantities employed in the simulation are
derived from experimental measurements, apart from the neutral particle data. 
Three cases are considered.
In the first case, the natural frequencies of tearing modes resonant in the plasma are 
determined by the ion flows at the corresponding rational surfaces, which is the prediction of nonlinear theory.
In the second case, the natural frequencies are 
determined by the guiding center flows at the rational surfaces.
 In the third case, the natural frequencies   are 
determined by the electron flows at the  rational surfaces, which is the prediction of linear theory.
The second case gives  the best agreement between the simulations and the  experimental observations.   The third case gives by far the worst agreement between the simulations 
and the  observations. The first case only leads to partial agreement between  the  simulations and the  observations. However, this lack of complete agreement
may be a consequence of using an inaccurate guess for the neutral particle distribution in the pedestal.  
\end{abstract}

@@

Figure~\ref{fig7} shows the widths of the regions of density and electron temperature flattening associated with the $n=3$ magnetic
island chains driven by the RMP in the pedestal region as functions of time. (See Sect.~III.F in Ref.~\onlinecite{rf2}.) (Note that the energy diffusivity shown in Fig.~\ref{fig3}  is used to calculate the
critical island width for electron temperature flattening, whereas the corresponding particle diffusivity is used to calculate
the critical island width for density flattening.\cite{flat}) It can be seen that the driven magnetic island chains are generally too narrow
to significantly flatten the density and electron temperature profiles,  implying that there is actually very little degradation of the pedestal energy and particle confinement
due to the applied RMP. 
However, in those situations in which the shielding of driven
magnetic reconnection due to plasma flow breaks down (i.e., the true natural frequency becomes zero), locked magnetic island chains are generated that are
 wide enough to flatten both the density and the electron temperature profiles.

 Ideally, in order to account for the
ELM-suppression windows in DIII-D discharge \#145380, according to the standard
hypothesis of how RMP-induced ELM-suppression operates, we require the major regions of density and temperature
flattening to correspond to the intersections of the yellow vertical bands and the horizontal dotted lines in Fig.~\ref{fig7}. If this is the case then
the density and temperature profiles are locally flattened at the top of the pedestal (which corresponds to
the horizontal dotted lines) during the ELM-suppression windows (which correspond to the yellow vertical bands). (Note that the top of the pedestal
is, somewhat arbitrarily, identified with the off-axis peak in the electron number density that is apparent in the top panel of Fig.~\ref{fig3}.) It can be seen
that the region of density and temperature flattening associated with the $m=11$ (brown) tearing mode driven by the RMP lines up fairly well with the
first ELM-suppression window. The region of density and temperature flattening associated with the $m=10$ (magenta) tearing mode 
lines up somewhat less well with the second ELM-suppression window. To be more exact, the region is located at the top of the pedestal,
but occurs slightly too early. However, the 
region of density and temperature flattening associated with the $m=9$ (cyan) tearing mode 
lines up very poorly with the third ELM-suppression window. To be more exact, the region is not really located at the top of the pedestal, and occurs
significantly too early. The ultimate reason for this poor alignment is that the major zero-crossing of the $m=9$ natural
frequency in the upper panel of Fig.~\ref{fig5} does not take place within the third ELM-suppression window (unlike the
major zero-crossings of the $m=11$ and $m=10$ natural frequencies, which do take place within the first and the second
ELM-suppression windows, respectively).

@@

\section*{Acknowledgements}
This research was funded by the U.S.\ Department of Energy under contract DE-FG02-04ER-54742.
The author would like to thank Q.M.~Hu and R.~Nazikian for providing the experimental data used in this paper. The
author would also like to thank J.-K.~Park and N.C.~Logan for providing guidance on how to run the {\tt GPEC} code.

\section*{Data Availabity Statement}
The data that support the findings of this study are available from the corresponding author upon reasonable request.

@@

Figure~\ref{fig14} also sketches the
typical plasma current and vorticity patterns in the nonlinear regime. Observe that there is now a considerable distinction between an X-point and an O-point. The reason for this is that the island is sufficiently wide that the plasma cannot diffuse across it resistively rapidly enough to
avoid being trapped inside the magnetic separatrix. Indeed, both the regions inside and outside the magnetic separatrix
are governed by a combination of flux-freezing and perturbed force balance, whereas the resistive
layer, within which the plasma can easily slip through the magnetic field,  is diverted onto the magnetic separatrix. Moreover, as
the island grows, the width of the separatrix layer becomes
increasing negligible with respect to the full island width. It
turns out that the separatrix layer does not significantly affect the evolution of a nonlinear magnetic island chain. Indeed,
the celebrated Rutherford island width evolution equation,
which governs the evolution of the island width, is simply a
$1\tfrac{1}{2}$-D evolution equation for a helical magnetic
equilibrium localized around the rational surface.\cite{ruth}
The Alfv\'{e}n time plays no role in this evolution (any more
than it plays a role in the $1\tfrac{1}{2}$-D evolution of
the global magnetic field). Likewise, the equation that
determines the phase velocity of a nonlinear magnetic island chain (see Ref.~\onlinecite{nl2}, and references therein) is essentially an
expression of the $1\tfrac{1}{2}$-D evolution of the plasma
flow profiles across the island region. As before, the  Alfv\'{e}n time plays no role in this evolution (any more
than it plays a role in the $1\tfrac{1}{2}$-D evolution of
the global plasma flow profiles).

@@

\begin{figure}
\includegraphics[height=7in]{fig3.pdf}
\caption{Top Panel: The red, green, blue, and cyan curves show the electron number density ($10^{19}\,{\rm m}^{-3}$),
electron temperature (keV), (thermal) ion temperature (keV), and
 C-VI ion number density  ($10^{18}\,{\rm m}^{-3}$)  profiles, respectively,  in  DIII-D discharge \#145380 at time $t=2500$ ms. Middle Panel:  
 ${\bf E}\times {\bf B}$ frequency profile in  DIII-D discharge \#145380 at time $t=2500$. Bottom Panel: The red, green, and blue curves show
 the perpendicular momentum, energy, and particle diffusivity profiles, respectively, 
 in  DIII-D discharge \#145380 at time $t=2500$ ms.
The   common vertical dotted lines indicate the location of the top
of the pedestal, ${\mit\Psi}_N=0.925$.} \label{fig3}
\end{figure}
